\input{thispreamble.tex}

\renewcommand\AUTHOR{jdoe5@cougars.ccis.edu} % CHANGE TO YOURS

\begin{document}
\topmattertwo

\input{preamble-instructions.tex}

\begin{center}
  \textsc{Honor Statement}
\end{center}
I, \answerbox{[Carl Alexander Dalebout]},
attest to the fact that the submitted work is my own and
is not the result of plagiarism.
Furthermore, I have not aided another student in the act of
plagiarism.

% ------------------------------------------------------------------------------
\begin{python}
from scoretable import *
\end{python}

% ------------------------------------------------------------------------------
\newpage
\input{preamble.tex}

% ------------------------------------------------------------------------------
\newpage
\nextq
\begin{itemize}

\item[(a)]
  ISA stands for
  
  \ANSWER\vspace{4mm}
  \begin{answercode}

  \end{answercode}

\item[(b)]
  CISC stands for
 
  \ANSWER\vspace{4mm}
  \begin{answercode}

  \end{answercode}

\item[(c)]
  RISC stands for
  \\
  \ANSWER\vspace{4mm}
  \begin{answercode}

  \end{answercode}
  
\item[(c)]
  \tf.
  An assembly instruction translates to one machine instruction.
  \\
  \ANSWER\vspace{4mm}
  \begin{answercode}
    t
  \end{answercode}
  
\item[(d)]
  The five parts of a computer are
  \\
  \ANSWER\vspace{4mm}
  \begin{answercode}
    input
    output
    Memory

  \end{answercode}

  
\item[(e)]
  The CPU or processor is made up of the following two parts:
  \\
  \ANSWER\vspace{4mm}
  \begin{answercode}
    Cache
    Chip
  \end{answercode}
  
\item[(f)]
  Which of the following provides the fastest
  storage mechanism. Choose one.
  \begin{itemize}
  \item[a.] CPU Cache
  \item[b.] Register file
  \item[b.] Memory
  \item[c.] External storage (such as hard drive)
  \end{itemize}
  \ANSWER\vspace{4mm}
  \begin{answercode}
    a
  \end{answercode}
\end{itemize}

% ------------------------------------------------------------------------------
\newpage
\nextq
Joe Cantcode wrote the following code to display the first
integer in the data segment 
to the
console.
But it doesn't work!!! Correct it.
In the code below, the first integer in the data segment is 2010.
But you cannot assume that it won't change in the future.
You must not change the data segment.

(From the management: This is a warm-up. Just insert one line.)

\textsc{Answer:}
\begin{answercode}  
	  .text
	  .globl    main
          
main:     
    la        $a0, odyssey
	  lw        $a0, 0($a0)
    li        $v0, 1
	  syscall

	  li        $v0, 10
	  syscall

          .data
odyssey:  .word 2010
\end{answercode}

% ------------------------------------------------------------------------------
\newpage
\nextq
Tom Toofaz always jumps into his project assignments and hack out code ... and
ends up spending his night in the office.
He's having problem implementing the following requirement
in MIPS with less than four \verb!t!--registers.
But his manager, Sloe Arnn Steady, told him that four is \textit{way} too many.
Help Tom and he'll buy you lunch.

Write a code segment in MIPS to compute
\verb!9 * a - (b + c) + d! where
\verb!$s0!, \verb!$s1!, \verb!$s2!, and
\verb!$s3! have values of the C/C++ integer variables
\verb!a!, \verb!b!, \verb!c!, \verb!d! respective.
The result must be placed in \verb!$s4!.
Your code must be as efficient as possible and you must use your registers
efficiently too.
You may assume that the \verb!t!-registers are not used elsewhere, and so you
can use the \verb!t!--registers in your code.
No such guarantee for the \verb!s!-registers -- therefore
you cannot change the values in
\verb!$s0!, \verb!$s1!, \verb!$s2!, and \verb!$s3!.
You also cannot use the \verb!a!-- and \verb!v!--registers.
Do not use the \verb!mul! or \verb!mult! command.
Write down the number of \verb!t!-registers used.
\\
\ANSWER
\begin{answercode}
# 9 * a - (b + c) + d
move  $t0, $s0       
add   $t0, $t0, $t0 
add   $t0, $t0, $t0 
add   $t0, $t0, $t0 # t0 = 8 * a
add   $t0, $t0, $s0 # t0 = 9 * a   
move  $t1, $s1      # t1 = b
add   $t1, $t1, $s2 # t1 = b + c
sub   $t0, $t0, $t1 # t0 = 9 * a - (b + c)
add   $s4, $t0, $s3 # s4 = 9 * a - (b + c) + d
\end{answercode}
\vspace{-4mm}
(Code fragment, not a complete program.)

Number of \verb!t!--registers used:
\\
\ANSWER
\begin{answercode}
2
\end{answercode}



% ------------------------------------------------------------------------------
\newpage
\nextq
The following is an attempt by Sue Longlunch to scan an array of 
words and to add 10 to each word in that array. 
The array has size 20 and base address is in register 
\verb!$s0!.
Your manager is fuming mad because there are bugs in the code
(and Sue is still at lunch) and have asked you to
do a code review and 
writing in some comments.
You should write \lq\lq \verb!should be: xxxxx!" to replace
the code or
\lq\lq\verb!insert here: xxxxx!" to insert some code between two lines.

\begin{answercode}

  ///// will come back to this
                                   COMMENTS TO SUEa
        addi    $t0, $s0, 20       should be: ????

loop:   blt     $s0, $t1, exit:    
                                   insert here: ????
        move    $t1, 0($s0)        

        addi    $s0, $s0, 1        

        addi    $t1, $t1, 10       
                                     
        jr      loop                 
exit:
\end{answercode}

% ------------------------------------------------------------------------------
\newpage
\nextq
While working on an artificial neural network,
your senior teammate Arnie Ned needs you to write a MIPS code fragment to 
\lq\lq clamp" the integer value in \verb!$s0! at \verb!-5! and \verb!5!.
In other words
if \verb!$s0! is less than \verb!-5!,
\verb!$s0! is set to \verb!-5!.
If the value of \verb!$s0! is
greater than \verb!5!,
\verb!$s0! is set to \verb!5!.
Otherwise, the value in \verb!$s0! is unchanged.

\textsc{Answer:}
\begin{answercode}
li  $t0, 5
ble $s0, $t0 clamp_upper # if(s0 <= t0) {/*do nothing*/}
  li  $s0, 5
clamp_upper:
li  $t0, -5
bge $s0, $t0 clamp_lower # if(s0 >= -5) {/*do nothing*/}
  li  $s0, 5
clamp_lower:
\end{answercode}
\vspace{-4mm}
(Code fragment, not a complete program.)

%------------------------------------------------------------------------------
\newpage
\nextq
What is the MIPS code fragment for the C code
\[
  \verb!f[i + 2] = f[i + 1] + f[i];!
\]
where \verb!f! is an array of integers.
The base
address of \verb!f! is associated with \verb!$t0!, \verb!i!
is associated with \verb!$t1!.

\textsc{Answer:}
\begin{answercode}
# t0 = f[]
# t1 = i
addu  $t2, $t1, $t1
addu  $t2, $t2, $t2
addu  $t2, $t2, $t0
lw    $t2, 0($t2)
addiu $t3, $t1, 1
addu  $t3, $t3, $t3
addu  $t3, $t3, $t3
addu  $t3, $t3, $t0
lw    $t3, 0($t3)
add   $t2, $t2, $t3 # t2 = f[i+1]+f[i]
addiu $t3, $t1, 2
addu  $t3, $t3, $t3 
addu  $t3, $t3, $t3 
addu  $t3, $t3, $t0 # t3 = f[i + 2]
sw    $t2, 0($t3)0

\end{answercode}

% ------------------------------------------------------------------------------
\newpage
\nextq
Your project leader Lee Pointa
wants you to write some MIPS code to perform the first pass of bubblesort
(in ascending order)
on an
integer array,
which will be used in a bubblesort that Bubba Short will use.
Lee says that the base address of the array is stored in \verb!$a0!
and the number of values in the array is stored in \verb!$a1!.
Another thing: Lee tells you he prefers you do not use an index variable, but
rather, use a pointer to iterate over the array.
(Look at his lastname.)

\textsc{Answer:}
\begin{answercode}
# a0 = x[]
# a1 = n
addu  $t0, $a1, $a1
addu  $t0, $t0, $t0 
addu  $t0, $t0, $a0 #  t0 = x[n]
bubblesort_loop:
blt   $a0, $t0, bubblesort_loop_end
  lw  $t1, 0($a0)
  lw  $t2, 4($a0)
  blt $t1, $t2, bubblesort_swap # if(t1 > t2) {}
    j bubblesort_loop_continue
  bubblesort_swap:
    sw  $t2, 0($a0)
    sw  $t1, 4($a0)
  bubblesort_loop_continue:
addiu $a0, $a0, 4
j   bubblesort_loop
bubblesort_loop_end:
\end{answercode}
\vspace{-4mm}
(Code fragment, not a complete program.)

%------------------------------------------------------------------------------
\newpage
\nextq
It's lunchtime ... the company quiz for today is below.
The winner is the first to answer all questions correctly.
Work fast: Lee Pointa is really good with these type of questions.

(a) What's in \verb!$s1!, \verb!$s2!, and \verb!$s3!
at the end of the code fragment (i.e., not complete program)?
(Write ERROR if the value is not an integer value.)
\begin{Verbatim}[frame=single,fontsize=\small]
        .text
        .globl main        
        [code omitted]
        la     $s0, L       # s0 = L
        lw     $s1, 8($s0)  # s1 = 44
        lw     $s2, 16($s0) # s2 = 46
        la     $s0, L1      # s0 = L1
        lw     $s3, -4($s0) # s3 = 49
        [code omitted]
        
        .data
L:      .word 42 43 44 45
        .word 46 47 48 49
L1:     .word 50 51 52 53
        .word 54 55 56 57
\end{Verbatim}
\textsc{Answer:}
\\
\verb!$s1! \answerbox{44} \hspace{1cm}
\verb!$s2! \answerbox{46} \hspace{1cm}
\verb!$s3! \answerbox{49} \\

(b)
What is the value of \verb!$s2! at the end of the code fragment?
(Write ERROR if the value is not an integer value.)
\begin{Verbatim}[frame=single,fontsize=\small]
        .text
        .globl main        
        [code omitted]
        la     $s0, W       
        lw     $s1, 4($s0)  # s1 = 2
        lw     $s2, 8($s1)  # error
        [code omitted]
        
        .data
X:      .word 1 2 3 4
Y:      .word 5 6 7 8
Z:      .word 9 10 11 12
W:      .word X Y Z
\end{Verbatim}
\textsc{Answer:} \\
\verb!$s2! \answerbox{ERROR}

%------------------------------------------------------------------------------
\newpage
\nextq
Write a MIPS code fragment that, starting with $n$ (assume stored in
\verb!$s0!),
it
continually computes $3n + 1$ if $n$ is odd or $n/2$
if $n$ is even
and store it back into $n$ until $n$ is $1$.
(\verb!/! is integer division.)
The values of $n$ are stores in an array in the data segment --
a label \verb!L! is already created for you (you don't have to create it).
For instance if the user
enters $3$, then the values are
\[
  3, 10, 5, 16, 8, 4, 2, 1
\]
For instance when $n = 3$, since $n$ is odd,
the next value of $n$ is $3n + 1 = 10$.
The next value for $n = 10$, since $n = 10$ is even,
is $n / 2 = 10 / 2 = 5$.
The values
\[
  3, 10, 5, 16, 8, 4, 2, 1
\]
are stored at the beginning of the data
segment.
Note that the starting value $n = 3$ is also stored.

\textsc{Answer:}
\begin{answercode}
  # s0 = n
  # t0 = index of L
  la    $t0, L
  sw    $s0, 0($t0)
  addiu $t0, $t0, 4
  loop:
  li    $t1, 1
  beq   $s0, $t1, loop_exit
  li    $t1, 2
  div   $s0, $t1
  hfhi  $t1
  beq   $t1, $0, comput_even # if(t1 == 0)
    mult  $s0, 3
    mflo  $s0
    addi  $s0, $s0, 1
    sw    $s0, 0($t0)
    addiu $t0, $t0, 4
    j   loop
  comput_even:
    div   $s0, $t1
    mfhi  $s0
    sw    $s0, 0($t0)
    addiu $t0, $t0, 4
  j loop
  loop_exit:

  .data
L:    .word
\end{answercode}

%------------------------------------------------------------------------------
\newpage
\nextq
Your boss just did a code review and found that Joe Cantcode
wrote some MIPS code implementing the following idea
in a critical part of the code to auto-correct
the tilt actuators on a high speed train:
\begin{Verbatim}[frame=single]
if (t0 == 0)
{
    t1 = 12;
}
else if (t0 == 1)
{
    t1 = 34;
}
else if (t0 == 2)
{
    t1 = 56;
}
else if (t0 == 3)
{
    t1 = 78;
}
\end{Verbatim}
where variables 
\verb!t0!,
\verb!t1!
and associated with registers
\verb!$t0!,
\verb!$t1!
respectively.
He has asked you rewrite the code to implement a switch 
in MIPS using the
jump table method 
\begin{Verbatim}[frame=single]
switch (t0)
{
    case 0:
        t1 = 12;
        break;
    case 1:
        t1 = 34;
        break;
    case 2:
        t1 = 56;
        break;
    case 3:
        t1 = 78;
}
\end{Verbatim}
so that he can do some 
performance analysis and see if the jump table method is faster.

\textsc{Answer.}
Implement the switch here
\begin{answercode}
        .text
        .globl main
main:
      li    $t0, 0           # your code must work for different values of t0
      la    $t1, Jtable
      addiu $t0, $t0, $t0
      addiu $t0, $t0, $t0
      addiu $t0, $t0, $t1
      jr    $t0 
      # implement switch below
      L0:
        li  $t1, 12
        j   exit
      L1:
        li  $t1, 34
        j   exit
      L2:
        li  $t1, 56
        j   exit
      L3:
        li  $t1, 78
      # At this point $t1 must be set to the right value based on your
      # implementation of switch
exit:
      li $v0, 10          # exit
      syscall

      .data
Jtable: L0, L1, L2, L3
\end{answercode}

%------------------------------------------------------------------------------
\newpage
\nextq
Implement the following C/C++ code fragment in
MIPS code fragment
\[
  \verb!y2 = x * x * x + 2 * x * x + 3 * x + 4;!
\]
where \verb!y2! is \verb!$s0! and \verb!x! is \verb!$s1!.
Assume that the cost of
multiplication is higher than the cost of addition.
Say the cost of multiplication is 20 and addition is 1.
You want your MIPS code to be as fast as possible
since the above is used in performing 
error correction codes in signal processing.
The project leader, Zig Nell,
claims that it can be done with two multiplications.

\textsc{Answer:}
\begin{answercode}
# s0 = y2
# s1 = X
addi  $t0, $s1, 2
mult  $t0, $s1
mflow $t0
addi  $t0, $t0, 3
mult  $t0, $s1
addi  $s0, $t0, 4
\end{answercode}

% ------------------------------------------------------------------------------
\newpage
\nextq
Just great.
Bubba Short is not back from lunch.
Lee Pointa said you have to write the bubblesort as a function  for him.
Remember that
the base address of the array is stored in \verb!$a0!
and the number of values in the array is stored in \verb!$a1!.
Just to be somewhat organized Lee told you to include at least
a \verb!swap! function that swaps the integer values at
addresses stored in \verb!$a0! and \verb!$a1!.
(You can add more functions if you like.
For instance you might want to have a function that performs
one pass of bubblesort.)

\textsc{Answer:}
\begin{answercode}
swap: # a0 = larger a1 = smaller
  lw  $t0, 0($a0)
  lw  $t1, 0($a1)
  sw  $t0, 0($a1)
  sw  $t1, 0($a0)
  jr  $ra

bubblesort_iterate:
  # a0 = x[]
  # a1 = i
  bubblesort_iterate_loop:
  blt   $a0, $a1, bubblesort_iterate_loop_end
    lw  $t1, 0($a0)
    lw  $t2, 4($a0)
    blt $t1, $t2, bubblesort_iterate_swap # if(t1 > t2) {}
      j bubblesort_iterate_loop_continue
    bubblesort_iterate_swap:
      sw  $t2, 0($a0)
      sw  $t1, 4($a0)
    bubblesort_iterate_loop_continue:
  addiu $a0, $a0, 4
  j   bubblesort_iterate_loop
  bubblesort_iterate_loop_end:
  jr    $ra

bubblesort: # a0 = x[], a1 = n
  addu  $t1, $a1, $a1
  addu  $t1, $t1, $t1
  addu  $t1, $t1, $a0 # t0 = x[n]
  move  $t0, $a0
  bubblesort_loop:
  beq   $t1, $t0, bubblesort_loop_exit
      addiu $sp, $sp, -12
      sw    $ra, 0($sp)
      sw    $a0, 4($sp)
      sw    $a1, 8($sp)
        move  $a1, $t1
        jal   bubblesort_iterate
      lw    $a1, 8($sp)
      lw    $a0, 4($sp)
      sw    $ra, 0($sp)
  addiu $t1, $t1, -4 
  j     bubblesort_loop
  bubblesort_loop_exit:
  jr    $ra
\end{answercode}

%------------------------------------------------------------------------------
\newpage
\input{instructions}
\end{document}
